Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-03 WSSSPE1_report/WSSSPE1_summary_report.tex
Added TODOs and refined Sections 1 to 3.

Because progress in scientific research is dependent on the quality and accessibility
of software at all levels, it is now critical to address many challenges related to the
development, deployment, and maintenance of reusable software. In addition, it is
essential that scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are already
acquiring some of these skills, and in particular a specialized class of software
developers is emerging in academic environments who are an integral and embedded
part of successful research teams. The WSSSPE workshop was intended provide a
forum for discussion of the challenges, including both positions and experiences.
The short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions are expected.
The level of interest in the workshop has led the organizers, working with some of the submitters
and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
the workshop website into a community website that can be used as a focus for further
discussion and progress.

@@

WSSSPE1 was organized in collaboration of the relatively small group of organizers and
a larger peer-review committee. This committee had early influence on the organization,
e.g., on the specific call for papers.

@@

\begin{itemize}
\item the development process that leads to new software
\begin{itemize}
\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}
\item the support and maintenance of existing software, including software engineering
\begin{itemize}
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}
\item education and training
\end{itemize}

@@

A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
reviewers to provide scores on relevance and comments to the organizers, which were used to 
decide which papers to associate with the workshop, and comments to the authors, which were 
provided back to the authors to allow them to improve their papers.

The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
that the reports were already self-published. The papers were also grouped into 3 areas, each of 
which was associated with a panel and discussion at the workshop.

@@

The workshop consisted of 2 keynote presentations and three panel/discussion sessions. The panels 
were organized based on a classification of the workshop submissions into three categories, following 
the themes of the call for papers as modified by the areas of the submissions. Representatives from 
each submission categories were appointed as panelists, to discuss the papers in that category.

@@

Philip is a basic biomedical scientist but has formed 4 software
companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

>>>>

Because progress in scientific research is dependent on the quality and accessibility
of software at all levels, it is now critical to address many challenges related to the
development, deployment, and maintenance of reusable software. In addition, it is
essential that scientists, researchers, and students are able to learn and adopt
software-related skills and methodologies. Established researchers are already
acquiring some of these skills, and in particular a specialized class of software
developers is emerging in academic environments who are an integral and embedded
part of successful research teams. The WSSSPE workshop was intended to provide a
forum for discussion of the challenges, including both positions and experiences.
The short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this journal publication.
Additional papers based on extended versions of workshop submissions are expected.
The level of interest in the workshop has led the organizers, working with some of the submitters
and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
the workshop website into a community website that can be used as a focus for further
discussion and progress.

\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July. It aims to further explore some of the key issues raised in WSSSPE. Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}

@@

WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
e.g., on the specific call for papers.

@@

\begin{itemize}
\item the development process that leads to new (versions of) software
\begin{itemize}
\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}
\item the support and maintenance of existing software, including software engineering
\begin{itemize}
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}
\item education and training
\end{itemize}

@@

A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
reviewers to provide scores on relevance and comments to the organizers and to the authors. The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.

The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
that the reports were already self-published. The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}. Each subject area was associated with a panel and discussion at the workshop.

@@

The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
were organized based on a classification of the workshop submissions into three categories, following 
the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
each submission categories were appointed as panelists, to discuss the papers in that category.

@@

Philip is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
